Synthesizing Quotient Lenses

### 2018-03-13 papers/QuotientLenses/main.tex
edits in progress of evaluation section; changes to macros for three different measures.

\newcommand{\QRESize}{\textbf{QS}}
\newcommand{\canonizeAndSpecSize}{\textbf{CS}}
\newcommand{\LensAndSpecSize}{\textbf{LS}}

@@

We have implemented QREs and the quotient lens synthesis algorithm described
above as an extension to the Boomerang
interpreter~\cite{boomerang,quotientlenses}. We have added QREs and QRE lens
synthesis tasks to Boomerang. Furthermore, these synthesis tasks evaluate to
Boomerang lens values, so Boomerang treats synthesized lenses the same as
hand-written ones. Intuition suggests that the increased expressiveness
of QREs should make many lens programming tasks easier, but we would like more
quantitative understanding of that claim. We thus conducted several experiments
that aim to provide answers to the following questions:

\begin{enumerate}
  \item Does synthesizing QRE lenses from QREs permit an easier development
  process than manually writing canonizers, and synthesizing the lenses between
  their canonized forms?  Does synthesizing QRE lenses from QREs permit an
  easier development process than writing the lenses by hand?

  \item Does our tool synthesize lenses fast enough to be used as part of a
  standard development process?
\end{enumerate}

All evaluations were performed on a 2.5 GHz Intel Core i7 processor with 16 GB
of 1600 MHz DDR3 running macOS High Sierra.

@@

To evaluate our QRE implementation, we adapted 39 lens synthesis tasks from the
benchmark suite of the original Optician system. These benchmarks are a
combination of custom benchmarks, benchmarks derived from
FlashFill~\cite{flashfill}, and benchmarks derived from Augeas~\cite{augeas2}.
In the pre-QRE version of Optician, many of these data formats had to be
modified to work with the bijectivity constraints that Optician required. For
instance, when one format permits whitespace where the other does not allow it,
the original versions of benchmark formats were modified to allow more
whitespace, thereby restoring bijectivity (but altering the data format). With
the new QRE support, we were able to remove these alterations when porting these
benchmarks. 10 of Optician's 39 benchmarks had formats modified in this way ---
introducing QREs removed these format alterations. (This experience alone
suggests that QREs make the lens development process easier.)

\subsection{Easing the Development Process}

@@

To get a sense of how much the new expressiveness that combining QRE lenses with
synthesis saves in terms of programmer effort, we compared the amount of code
that would need to be written for each of the lens benchmarks. We use number of
nodes in the abstract syntax tree as a measure of code size, and calculated
three quantities for each lens synthesis benchmark task. We only compared the
sizes on benchmarks where the canonization functions are nontrivial.
%
\begin{itemize}
  \item[\QRESize{}] the number of AST nodes in the QRE
  specifications (including examples).
  \item[\canonizeAndSpecSize{}]  the number of
  AST nodes in $W(q_1)$ summed with the number of AST nodes in $\canonize(q_1)$
  and the number of examples.
  \item[\LensAndSpecSize{}] the number of AST nodes
  in $W(q_1)$ summed with the number of AST nodes in the synthesized QRE lens.
\end{itemize}

The value of \QRESize{} is our measure of the programmer burden for
writing our benchmarks---it measures the number of AST nodes the programmer has
to write using the full expressiveness of QREs and synthesis.

The value of \canonizeAndSpecSize{} estimates the burden for writing our
benchmarks within the Optician system, without QREs but still using
synthesis. (This is only an approximation, as both $W(q_1)$ and $\canonize(q_1)$ are
automatically generated from the QRE, and it is possible that a human-written
version might be smaller.)

The value of \LensAndSpecSize{} estimates the burden in writing our benchmarks
in Boomerang, using neither QREs nor synthesis. (This is also an
approximation, as $W(q_1)$ and $\canonize(q_1)$ and the bijective lens are all
automatically generated.)

>>>>

\newcommand{\QRESize}{\textbf{QS}}
\newcommand{\canonizeAndSpecSize}{\textbf{BS}}
\newcommand{\LensAndSpecSize}{\textbf{NS}}

@@

We have implemented QREs and the quotient lens synthesis algorithm described
above as an extension to the Boomerang
interpreter~\cite{boomerang,quotientlenses}. 
The synthesis algorithm produces Boomerang lens values, so Boomerang
treats synthesized lenses the same as hand-written ones.
To evaluate the effectiveness of QREs, QRE lenses, and QRE lens
synthesis, we conducted experiments to answer the following questions:
%
\begin{itemize}
  \item {\bf Ease of use.} Does synthesizing QRE lenses from QREs permit an
  easier development process than writing lenses by hand?
  Does synthesizing QRE lenses from QREs permit an easier development
  process than manually writing canonizers and then synthesizing lenses between
  their canonized forms?  
  \item {\bf Performance.} Is the synthesis algorithm/implementation fast enough to be
  used as part of a standard development process?
\end{itemize}
%
All evaluations were performed on a 2.5 GHz Intel Core i7 processor with 16 GB
of 1600 MHz DDR3 running macOS High Sierra.

@@

To evaluate our QRE implementation, we adapted 39 lens synthesis tasks
from the benchmark suite of the original Optician system. These
benchmarks are a combination of custom benchmarks, benchmarks derived
from FlashFill~\cite{flashfill}, and benchmarks derived from
Augeas~\cite{augeas2}.  In the pre-QRE version of Optician, 10 of
these data formats had to be modified to work with the bijectivity
constraints that Optician required. For instance, when one
representation permits whitespace where the other does not, we
modified the original version of the benchmark to allow more whitespace,
thereby restoring bijectivity (but altering the data format). With the
new QRE support, we were able to remove these alterations.  This
experience alone suggests that QREs make the lens development process
more flexible.

\subsection{Ease of Use}

@@

To evaluate the impact of QRE lens synthsis on programmer effort,
we focus our attention on the 10 problems in the benchmark suite that
are not bijective and hence require non-trivial canonizers. 
(Optician already handles the other problems with minimal programmer
effort.)

We are interested in comparing three different approaches, which vary
in the amount of synthesis used. 
In the first approach, which we call \QRESize{} for QRE Synthesis, the programmer uses
QRE synthesis.  She must write QRE specifications of the source and
target formats and she may give examples.
In the second approach, which we call \canonizeAndSpecSize{} for
Bijective Synthesis, the
programmer uses bijective lens synthesis \`a la the original version
of Optician.  She must write canonizers by hand, along with the
regular expressions describing the external
representations of the source and target formats. (The internal
regular expressions can be inferred from the canonizers.) She may also
provide examples to help in the synthesis of the bijective lens.
In the third approach, which we call \LensAndSpecSize{} for No Synthesis, the
programmer writes the lens between the source and target lenses
entirely by hand.

To measure the programmer effort using each of these approaches, we
calculate three different quantities for each problem in the benchmark suite:
%
\begin{itemize}
  \item[\QRESize{}:] 
  The number of AST nodes in the QRE specifications for the source and
  target formats, including examples. 
  \item[\canonizeAndSpecSize{}:] 
  The sum of (1) the number of AST nodes in $W(q)$ for each QRE $q$ in the source and target
  formats, (2) the number of AST nodes $\canonize(q)$ for each such QRE $q$ with a
  non-trivial canonizer, and (3) the number of AST nodes in the
  examples.  We use (1) and (2) to estimate the burden of describing
  the external source and target formats and of writing the requisite canonizers
  by hand.  We count the nodes in the examples because they would be
  fed to the bijective synthesizer.  
  These counts are an approximation, as both $W(q)$ and $\canonize(q)$ are
  automatically generated from the QRE, and it is possible that a human-written
  version might be smaller.
  \item[\LensAndSpecSize{}:] The sum of (1) the number of AST nodes in
  $W(q)$ for each QRE $q$ in the source and target formats and (2) the
  number of AST nodes in the synthesized QRE lens.  We use (1) to
  estimate the burdern for describing the source and target formats
  and (2) to estimate the burdern for writing the appropriate lens by
  hand. These counts are also approximations, as
  $W(q)$ and the synthesized lens may be larger than one written by hand.
\end{itemize}
